% ==============================================================
% Number Theory (Discrete)
% ==============================================================

\section{Number Theory (Discrete)}

\subsection{GCD, Bezout, extended Euclidean}

\begin{remark}
Throughout, we write $a \mid b$ to mean ``$a$ divides $b$'', i.e.\ $\exists q \in \mathbb{Z} : aq = b$. For $M \in \mathbb{Z}^+$, we define $\mathbb{Z}_M := \{1, 2, \ldots, M-1\}$. A key elementary rule used throughout: if $a \equiv a' \pmod{M}$ and $b \equiv b' \pmod{M}$, then $ab \equiv a'b' \pmod{M}$.
\end{remark}

\subsubsection*{Greatest Common Divisor and Coprimality}

\begin{definition}
Let $a, b$ be integers, not both zero. The \textbf{greatest common divisor} $\gcd(a,b)$ is the largest integer $d$ such that $d \mid a$ and $d \mid b$.
\end{definition}

\begin{definition}
Two integers $a$ and $b$ are \textbf{coprime} (or \emph{relatively prime}) if $\gcd(a,b) = 1$.
\end{definition}

\begin{example}
The divisors of 15 are $\{1,3,5,15\}$ and the divisors of 28 are $\{1,2,4,7,14,28\}$. Their only common divisor is 1, so $\gcd(15,28)=1$ and they are coprime. By contrast, $\gcd(15,25)=5$, so 15 and 25 are not coprime.
\end{example}

\subsubsection*{Bézout's Identity}

\begin{theorem}[Bézout's Identity]
Let $a$ and $b$ be integers, not both zero. There exist integers $x$ and $y$ (called \textbf{Bézout's coefficients}) such that
\[
ax + by = \gcd(a,b).
\]
\end{theorem}

\begin{example}
Since $\gcd(15,28)=1$, there exist $x,y \in \mathbb{Z}$ with $15x + 28y = 1$. One solution is $x=15,\, y=-8$. Note that solutions are not unique: for any $k \in \mathbb{Z}$, replacing $x$ by $x + k\cdot 28$ and $y$ by $y - k \cdot 15$ gives another solution.
\end{example}

\begin{remark}
The equation $ax + by = c$ has integer solutions if and only if $\gcd(a,b) \mid c$. If $d = \gcd(a,b) \nmid c$, then $d$ divides the left-hand side but not the right, giving a contradiction.
\end{remark}

\subsubsection*{Extended Euclidean Algorithm}

The \textbf{Extended Euclidean Algorithm} (EEA) simultaneously computes $\gcd(a,b)$ and Bézout's coefficients by tracking three sequences $(r_i), (s_i), (t_i)$ satisfying $s_i a + t_i b = r_i$ at each step.

\medskip
\noindent\textbf{Algorithm} (Input: $a \geq b \geq 0$; Output: $g, x, y$ with $g = \gcd(a,b)$ and $ax+by=g$):
\begin{enumerate}
    \item Initialise: $r_0=a,\; r_1=b;\quad s_0=1,\; s_1=0;\quad t_0=0,\; t_1=1$.
    \item While $r_i > 0$: compute quotient $q_i = \lfloor r_{i-1}/r_i \rfloor$, then
    \[
    r_{i+1} = r_{i-1} - q_i r_i, \quad s_{i+1} = s_{i-1} - q_i s_i, \quad t_{i+1} = t_{i-1} - q_i t_i.
    \]
    \item Return $r_{i-1},\, s_{i-1},\, t_{i-1}$.
\end{enumerate}

\begin{remark}
The while loop executes $O(\log b)$ times. Accounting for integer multiplication, the total time complexity is $O(\log^2 a)$.
\end{remark}

\begin{example}
Running EEA on $(a,b) = (28,15)$ yields $\gcd(28,15)=1$ with Bézout coefficients $x=7,\; y=-13$, i.e.\ $28 \cdot 7 + 15 \cdot (-13) = 196 - 195 = 1$. (Compare with $x=15, y=-8$ for the pair $(15,28)$ — coefficients depend on input order and are not unique.)
\end{example}

\subsubsection*{Linear Diophantine Equations}

\begin{definition}
A \textbf{linear Diophantine equation} is an equation of the form $ax + by = c$ where $a, b, c \in \mathbb{Z}$ and integer solutions $(x, y)$ are sought.
\end{definition}

\begin{theorem}[Existence of Integer Solutions]
The equation $ax + by = c$ has integer solutions if and only if $\gcd(a,b) \mid c$.
\end{theorem}
\begin{proof}
Let $d = \gcd(a,b)$. Since $d \mid a$ and $d \mid b$, we have $d \mid (ax + by)$ for any integers $x, y$. So if $ax + by = c$, then $d \mid c$, which is necessary.

Conversely, if $d \mid c$, write $c = d \cdot m$. By Bézout's Identity there exist $x_0, y_0$ with $ax_0 + by_0 = d$. Multiplying through by $m$ gives $a(mx_0) + b(my_0) = dm = c$, so $(x,y) = (mx_0, my_0)$ is a solution.
\end{proof}

\begin{remark}[General Solution]
If $(x_0, y_0)$ is one particular solution to $ax + by = c$, then the complete set of integer solutions is:
\[
x = x_0 + \frac{b}{d}\,k, \qquad y = y_0 - \frac{a}{d}\,k, \qquad k \in \mathbb{Z},\quad d = \gcd(a,b).
\]
\end{remark}

\begin{example}[EEA with back-substitution: $\gcd(101, 23)$]
Euclidean algorithm:
\[
101 = 4(23) + 9,\quad 23 = 2(9) + 5,\quad 9 = 1(5) + 4,\quad 5 = 1(4) + 1,\quad 4 = 4(1).
\]
So $\gcd(101,23) = 1$. Back-substituting to express 1 as $101x + 23y$:
\[
1 = 5 - 1(4) = 5 - (9-5) = 2(5) - 9 = 2(23-2\cdot9) - 9 = 2(23) - 5(9) = 2(23) - 5(101 - 4\cdot23)
\]
\[
= 22(23) - 5(101).
\]
Hence one Bézout pair is $(x, y) = (-5,\, 22)$, verifying $101(-5) + 23(22) = -505 + 506 = 1$.
\end{example}

\begin{example}[Solving $18x + 30y = 12$]
Check: $\gcd(18, 30) = 6$ and $6 \mid 12$, so integer solutions exist.

Divide through by 6: $3x + 5y = 2$. Try $y = 1$: $3x = -3$, so $x = -1$.

One solution is $(x,y) = (-1,\, 1)$. Check: $18(-1) + 30(1) = -18 + 30 = 12$. \checkmark

The general solution is $x = -1 + 5k$, $y = 1 - 3k$ for $k \in \mathbb{Z}$.
\end{example}

\begin{example}[$\gcd(84,30)$ via back-substitution]
Euclidean algorithm: $84 = 2(30) + 24$,\; $30 = 1(24) + 6$,\; $24 = 4(6)$. So $\gcd(84,30)=6$.

Since $\gcd(84,30) = 6 \neq 1$, they are \textbf{not coprime}.

Back-substituting: $6 = 30 - 24 = 30 - (84 - 2\cdot30) = 3(30) - 84$, giving $(x,y) = (-1,\, 3)$. Check: $84(-1) + 30(3) = -84 + 90 = 6$. \checkmark
\end{example}

\subsection{Congruences and modular inverses}

\subsubsection*{Residue Classes and Congruence}

\begin{definition}
For $M \in \mathbb{Z}^+$, integers $a$ and $b$ are \textbf{congruent modulo $M$}, written $a \equiv b \pmod{M}$, if $M \mid (a - b)$. The \textbf{residue class} of $a$ modulo $M$ is the set of all integers congruent to $a$:
\[
[a]_M = \{ a + kM : k \in \mathbb{Z} \}.
\]
The set of all residue classes modulo $M$ is denoted $\mathbb{Z}/M\mathbb{Z}$.
\end{definition}

\begin{theorem}[Congruence Multiplication]
If $a \equiv a' \pmod{M}$ and $b \equiv b' \pmod{M}$, then $ab \equiv a'b' \pmod{M}$.
\end{theorem}
\begin{proof}
Since $M \mid (a - a')$ and $M \mid (b - b')$, write $ab - a'b' = a(b - b') + b'(a - a')$. Then $M$ divides both terms on the right, hence $M \mid (ab - a'b')$, giving $ab \equiv a'b' \pmod{M}$.
\end{proof}

\begin{remark}
This rule is what makes it valid to reduce intermediate results during power computations: at each step, replace any number by its remainder mod $M$ before multiplying again.
\end{remark}

\subsubsection*{The Multiplicative Group $\mathbb{Z}_M^*$}

\begin{definition}
Let $M$ be a positive integer. The \textbf{multiplicative group of integers modulo $M$} is
\[
\mathbb{Z}_M^* = \{ a \in \mathbb{Z}_M \mid \gcd(a, M) = 1 \},
\]
equipped with the binary operation of multiplication modulo $M$.
\end{definition}

\begin{lemma}
$\mathbb{Z}_M^*$ is a commutative group under multiplication modulo $M$.
\end{lemma}
\begin{proof}[Proof sketch]
\textit{Closure:} if $\gcd(a,M)=1$ and $\gcd(b,M)=1$ then $\gcd(ab,M)=1$.
\textit{Associativity:} inherited from integer multiplication.
\textit{Identity:} the element $1 \in \mathbb{Z}_M^*$ satisfies $\gcd(1,M)=1$.
\textit{Inverses:} for any $a \in \mathbb{Z}_M^*$, since $\gcd(a,M)=1$, Bézout's identity gives $x,y$ with $ax + My = 1$. Reducing modulo $M$ yields $ax \equiv 1 \pmod{M}$, so $x \bmod M$ is the multiplicative inverse of $a$.
\end{proof}

\begin{example}
For $M=15$: $\mathbb{Z}_{15}^* = \{1,2,4,7,8,11,13,14\}$. Since $2 \times 8 = 16 \equiv 1 \pmod{15}$, the elements 2 and 8 are mutual inverses in this group.
\end{example}

\begin{example}[$\mathbb{Z}_{10}^*$ and modular inverses]
The elements of $\mathbb{Z}_{10}^*$ are those $a \in \{1,\ldots,9\}$ with $\gcd(a,10)=1$:
\[
\mathbb{Z}_{10}^* = \{1,\, 3,\, 7,\, 9\}.
\]
To find $3^{-1} \pmod{10}$: we need $3x \equiv 1 \pmod{10}$. Testing: $3 \times 7 = 21 \equiv 1 \pmod{10}$, so $3^{-1} \equiv 7$.

\textit{Alternative via Euler:} $\varphi(10)=4$, so $3^{-1} \equiv 3^{\varphi(10)-1} = 3^3 = 27 \equiv 7 \pmod{10}$. Same answer.

To find $7^{-1} \pmod{15}$: test $7 \times 13 = 91 = 90 + 1 \equiv 1 \pmod{15}$, so $7^{-1} \equiv 13$.
\end{example}

\subsubsection*{Euler's Totient Function}

\begin{definition}
For a positive integer $M$, \textbf{Euler's totient function} $\varphi(M)$ counts the integers in $\{1,\ldots,M\}$ that are coprime to $M$. Equivalently, $\varphi(M) = |\mathbb{Z}_M^*|$.
\end{definition}

\begin{theorem}
If $M$ is prime, then $\varphi(M) = M - 1$.
\end{theorem}
\begin{proof}
A prime $M$ has only divisors 1 and $M$, so every $k \in \mathbb{Z}_M$ satisfies $\gcd(k,M)=1$. Thus $\varphi(M) = |\mathbb{Z}_M| = M-1$.
\end{proof}

\begin{theorem}
If $p$ and $q$ are distinct primes and $M = pq$, then $\varphi(M) = (p-1)(q-1)$.
\end{theorem}
\begin{proof}
The integers in $\{1,\ldots,M\}$ that share a factor with $pq$ are exactly the multiples of $p$ and the multiples of $q$ (by primality). There are $q$ multiples of $p$ and $p$ multiples of $q$ in that range; $pq$ itself is counted twice. So the count of integers \emph{not} coprime to $pq$ is $p + q - 1$, giving
\[
\varphi(pq) = pq - (p+q-1) = (p-1)(q-1). \qedhere
\]
\end{proof}

\begin{theorem}[General Totient Formula]
For any $M \in \mathbb{Z}^+$ with prime factorisation $M = p_1^{k_1} p_2^{k_2} \cdots p_r^{k_r}$,
\[
\varphi(M) = M \prod_{p \mid M} \left(1 - \frac{1}{p}\right).
\]
\end{theorem}

\begin{example}[Worked totient computations]
\begin{itemize}
    \item $\varphi(13) = 13 - 1 = 12$ \quad (13 is prime).
    \item $\varphi(21) = \varphi(3 \cdot 7) = (3-1)(7-1) = 12$.
    \item $\varphi(15) = \varphi(3 \cdot 5) = (3-1)(5-1) = 8$.
    \item $\varphi(10) = \varphi(2 \cdot 5) = (2-1)(5-1) = 4$.
\end{itemize}
\end{example}

\begin{remark}
A useful identity: $\displaystyle\sum_{d \mid M} \varphi(d) = M$, provable by partitioning the fractions $\tfrac{1}{M}, \tfrac{2}{M}, \ldots, \tfrac{M}{M}$ by the denominator of their reduced form.
\end{remark}

\subsubsection*{Summatory Totient}

\begin{definition}[Summatory totient]
$\displaystyle\Phi(n) = \sum_{k=1}^{n} \varphi(k)$.
\end{definition}

\begin{remark}[Combinatorial meaning]
$\Phi(n)$ counts the total number of reduced proper fractions with denominator at most $n$, since $\varphi(k)$ counts valid numerators for denominator $k$. (See also the counting section below.)
\end{remark}

\begin{theorem}[Recursive formula for $\Phi(n)$]
\[
\Phi(n) = \frac{n(n+1)}{2} - \sum_{\ell=2}^{n} \Phi\!\left(\left\lfloor \tfrac{n}{\ell} \right\rfloor\right).
\]
\end{theorem}
\begin{proof}
Start from the identity $\sum_{d \mid k} \varphi(d) = k$. Sum over $k = 1, \ldots, n$:
\[
\sum_{k=1}^n k = \sum_{k=1}^n \sum_{d \mid k} \varphi(d) = \sum_{d=1}^n \varphi(d) \cdot \left\lfloor \tfrac{n}{d} \right\rfloor.
\]
Isolate the $d=1$ term ($\varphi(1)=1$) and rearrange:
\[
\Phi(n) = \frac{n(n+1)}{2} - \sum_{\ell=2}^{n} \left\lfloor \tfrac{n}{\ell} \right\rfloor \varphi(\ell) \quad \Longrightarrow \quad \text{(after re-indexing)} \quad \Phi(n) = \frac{n(n+1)}{2} - \sum_{\ell=2}^{n} \Phi\!\left(\left\lfloor \tfrac{n}{\ell} \right\rfloor\right).
\]
\end{proof}

\begin{remark}[Floor-grouping trick for fast evaluation]
For a fixed $n$, many consecutive values of $\ell$ share the same $\lfloor n/\ell \rfloor$. Specifically, for each distinct quotient $v = \lfloor n/\ell \rfloor$, the range of $\ell$ achieving it is $[\lfloor n/(v+1) \rfloor + 1,\; \lfloor n/v \rfloor]$. Iterating over blocks of equal quotients reduces the sum to $O(\sqrt{n})$ distinct terms. Combined with memoisation of $\Phi(\lfloor n/\ell \rfloor)$, this gives an $O(n^{2/3})$ algorithm.
\end{remark}

\subsubsection*{Euler's Theorem}

\begin{theorem}[Euler's Theorem]
If $a, M \in \mathbb{Z}^+$ are coprime, then
\[
a^{\varphi(M)} \equiv 1 \pmod{M}.
\]
\end{theorem}
\begin{proof}
Let $\mathbb{Z}_M^* = \{x_1, \ldots, x_k\}$ with $k = \varphi(M)$. Since $a \in \mathbb{Z}_M^*$, multiplication by $a$ permutes the elements of the group (as $a$ is invertible). Therefore
\[
\prod_{i=1}^k x_i \;\equiv\; \prod_{i=1}^k (ax_i) \;=\; a^k \prod_{i=1}^k x_i \pmod{M}.
\]
Since each $x_i$ is coprime to $M$, their product $P = \prod x_i$ is also coprime to $M$ and thus invertible modulo $M$. Multiplying both sides by $P^{-1}$ gives $1 \equiv a^k = a^{\varphi(M)} \pmod{M}$.
\end{proof}

\begin{corollary}[Fermat's Little Theorem]
If $p$ is prime and $\gcd(a,p)=1$, then $a^{p-1} \equiv 1 \pmod{p}$.
\end{corollary}

\begin{remark}[Multiplicative Inverse via Euler's Theorem]
Since $a^{\varphi(M)} \equiv 1 \pmod{M}$, we can factor this as $a \cdot a^{\varphi(M)-1} \equiv 1$, so:
\[
a^{-1} \equiv a^{\varphi(M)-1} \pmod{M} \qquad \text{(whenever } \gcd(a,M)=1\text{)}.
\]
This gives a formula for the inverse using only the totient. For example, $2^{-1} \pmod{15}$: since $\varphi(15)=8$, we get $2^{-1} \equiv 2^7 = 128 \equiv 8 \pmod{15}$. Check: $2 \times 8 = 16 \equiv 1$. \checkmark

\textbf{Warning:} If $\gcd(a,M) \neq 1$, then $a \notin \mathbb{Z}_M^*$, Euler's theorem does not apply, and $a$ has \emph{no} multiplicative inverse modulo $M$. For instance, $\gcd(3,15)=3 \neq 1$, so 3 has no inverse mod 15.
\end{remark}

\begin{example}[Euler's theorem verification: $a=2$, $M=15$]
Check: $\gcd(2,15)=1$ and $\varphi(15)=8$. Then $2^8 = 256 = 15 \cdot 17 + 1$, so $2^8 \equiv 1 \pmod{15}$. \checkmark
\end{example}

\begin{example}[Large power: $3^{100} \pmod{10}$]
Since $\gcd(3,10)=1$, Euler's theorem applies with $\varphi(10)=4$, giving $3^4 \equiv 1 \pmod{10}$.
Write $100 = 4 \times 25$, so $3^{100} = (3^4)^{25} \equiv 1^{25} = 1 \pmod{10}$.

The last digit of $3^{100}$ is \textbf{1}.
\end{example}

\subsection{RSA and applications to cryptography}

\subsubsection*{Order of an Element}

\begin{definition}
Let $a \in \mathbb{Z}_M^*$. The \textbf{order} of $a$ modulo $M$, written $\mathrm{ord}_M(a)$, is the smallest positive integer $r$ such that
\[
a^r \equiv 1 \pmod{M}.
\]
\end{definition}

\begin{remark}[Computing the order: repeated multiplication mod $M$]
To compute $\mathrm{ord}_M(a)$, evaluate $a^1, a^2, a^3, \ldots \pmod{M}$ using the congruence rule: at each step, multiply the \emph{current reduced value} by $a$, then reduce modulo $M$. Stop at the first exponent giving 1.

This is efficient because the congruence multiplication property lets you keep numbers small throughout.
\end{remark}

\begin{example}[Orders in $\mathbb{Z}_7^*$]
$\mathbb{Z}_7^* = \{1,2,3,4,5,6\}$, with $\varphi(7)=6$. So all orders divide 6, and the only possible orders are $1, 2, 3, 6$.

\medskip
\noindent\textit{Order of 2:}
$2^1\equiv2,\; 2^2\equiv4,\; 2^3=8\equiv1 \pmod{7}$.
So $\mathrm{ord}_7(2) = 3$.

\medskip
\noindent\textit{Order of 3:}
\[
3^1\equiv3,\quad 3^2=9\equiv2,\quad 3^3\equiv3\cdot2=6,\quad 3^4\equiv3\cdot6=18\equiv4,\quad 3^5\equiv3\cdot4=12\equiv5,\quad 3^6\equiv3\cdot5=15\equiv1.
\]
So $\mathrm{ord}_7(3) = 6$. Since $|\mathbb{Z}_7^*|=6$, the element 3 is a \textbf{generator} (primitive root mod 7).

\medskip
\noindent\textit{Check order divides totient:} $\mathrm{ord}_7(2) = 3$ divides $\varphi(7) = 6$. \checkmark
\end{example}

\begin{lemma}[Order Divides Totient]
For any $a \in \mathbb{Z}_M^*$, we have $\mathrm{ord}_M(a) \mid \varphi(M)$.
\end{lemma}
\begin{proof}
Let $r = \mathrm{ord}_M(a)$. By Euler's Theorem, $a^{\varphi(M)} \equiv 1$. Write $\varphi(M) = qr + s$ with $0 \le s < r$. Then
\[
1 \equiv a^{\varphi(M)} = a^{qr+s} = (a^r)^q \cdot a^s \equiv 1^q \cdot a^s = a^s \pmod{M}.
\]
Since $r$ is the \emph{smallest} positive exponent giving 1, we need $s = 0$, so $r \mid \varphi(M)$.
\end{proof}

\begin{corollary}
For a prime $p$, $\mathrm{ord}_p(a)$ divides $p-1$ for every $a \in \mathbb{Z}_p^*$.
\end{corollary}

\subsubsection*{Cyclic Groups and Generators}

\begin{definition}
A commutative group $(G, \cdot)$ is \textbf{cyclic} if there exists $g \in G$ (a \textbf{generator}, written $G = \langle g \rangle$) such that every element of $G$ is a power of $g$. In $\mathbb{Z}_M^*$, a generator is called a \textbf{primitive root modulo $M$}.
\end{definition}

\begin{example}
In $\mathbb{Z}_5^* = \{1,2,3,4\}$, the powers of 2 are:
\[
2^1 \equiv 2,\quad 2^2 \equiv 4,\quad 2^3 \equiv 3,\quad 2^4 \equiv 1 \pmod{5}.
\]
Since these cover all elements, 2 is a generator and $\mathbb{Z}_5^* = \langle 2 \rangle$ is cyclic.
\end{example}

\begin{remark}[Possible orders and generators in $\mathbb{Z}_p^*$]
For prime $p$:
\begin{itemize}
    \item $|\mathbb{Z}_p^*| = \varphi(p) = p-1$.
    \item Every element order divides $p-1$ (Order Divides Totient).
    \item Therefore the \textbf{only possible orders} are the positive divisors of $p-1$.
    \item A \textbf{generator} must produce all $p-1$ elements, so it must have order exactly $p-1$.
\end{itemize}
\end{remark}

\begin{example}[All orders in $\mathbb{Z}_{11}^*$]
$p=11$, $\varphi(11)=10$, possible orders $= \{1, 2, 5, 10\}$ (divisors of 10).

\medskip
\begin{center}
\begin{tabular}{c|l|c}
Element $a$ & Powers mod 11 (until 1) & $\mathrm{ord}_{11}(a)$ \\
\hline
1 & $1$ & 1 \\
10 & $10,\;1$ & 2 \\
3 & $3,9,5,4,1$ & 5 \\
4 & $4,5,9,3,1$ & 5 \\
5 & $5,3,4,9,1$ & 5 \\
9 & $9,4,3,5,1$ & 5 \\
2 & $2,4,8,5,10,9,7,3,6,1$ & 10 \\
6 & $6,3,7,9,10,5,8,4,2,1$ & 10 \\
7 & $7,5,2,3,10,4,6,9,8,1$ & 10 \\
8 & $8,9,6,4,10,3,2,7,5,1$ & 10 \\
\end{tabular}
\end{center}

\medskip
Elements with order 10 (= $p-1$) are the generators: $\{2, 6, 7, 8\}$. Since 2 produces all 10 elements, $\mathbb{Z}_{11}^* = \langle 2 \rangle$ and 2 is a primitive root mod 11.
\end{example}

\begin{theorem}[Existence of Primitive Roots]
For any prime $p$, the group $\mathbb{Z}_p^*$ is cyclic; i.e.\ there exists a primitive root modulo $p$.
\end{theorem}
\begin{proof}
For each divisor $d$ of $p-1$, let $N_p(d)$ denote the number of elements in $\mathbb{Z}_p^*$ of order exactly $d$.

\textit{Step 1.} We show $N_p(d) \in \{0, \varphi(d)\}$. Suppose $N_p(d) > 0$, so some $a \in \mathbb{Z}_p^*$ has order $d$. The elements $1, a, a^2, \ldots, a^{d-1}$ are $d$ distinct roots of $x^d - 1 \equiv 0 \pmod p$. Since a degree-$d$ polynomial over a prime field has at most $d$ roots, these are \emph{all} the roots. An element $a^k$ has order $d/\gcd(k,d)$; it has order exactly $d$ iff $\gcd(k,d)=1$. There are $\varphi(d)$ such $k$, so $N_p(d) = \varphi(d)$.

\textit{Step 2.} Every element of $\mathbb{Z}_p^*$ has some order dividing $p-1$ (by the corollary above), and orders partition the $p-1$ elements:
\[
\sum_{d \mid p-1} N_p(d) = p-1.
\]

\textit{Step 3.} From step 1, $N_p(d) \le \varphi(d)$ for all $d$. Combined with the identity $\sum_{d \mid p-1} \varphi(d) = p-1$ (see the remark in the previous subsection), both sides of $\sum N_p(d) \le \sum \varphi(d)$ equal $p-1$. Therefore $N_p(d) = \varphi(d)$ for every divisor $d$.

In particular, $N_p(p-1) = \varphi(p-1) \ge 1$, so at least one element has order $p-1$ and generates $\mathbb{Z}_p^*$.
\end{proof}

\begin{remark}[Relevance to Shor's Algorithm]
The order-finding problem — given $a$ and $M$, find $\mathrm{ord}_M(a)$ — is computationally hard classically but can be solved efficiently on a quantum computer. Shor's algorithm reduces integer factorisation to order-finding in $\mathbb{Z}_N^*$: if $r = \mathrm{ord}_N(a)$ is even and $a^{r/2} \not\equiv -1 \pmod{N}$, then $\gcd(a^{r/2} \pm 1, N)$ yields a non-trivial factor of $N$, breaking RSA.
\end{remark}

\subsection{Counting Reduced Proper Fractions}

\begin{definition}
A fraction $n/d$ is a \textbf{reduced proper fraction} if $1 \le n < d$ and $\gcd(n,d)=1$.
\end{definition}

\begin{theorem}
The total number of reduced proper fractions with denominator $d \le N$ is
\[
\sum_{d=2}^{N} \varphi(d).
\]
\end{theorem}
\begin{proof}
For a fixed denominator $d$, the valid numerators are $n \in \{1, 2, \ldots, d-1\}$ with $\gcd(n,d)=1$. By definition of Euler's totient, there are exactly $\varphi(d)$ such numerators. Summing over all $d$ gives the result.
\end{proof}

\begin{example}
For $d \le 1{,}000{,}000$, the total number of reduced proper fractions is
\[
\sum_{d=2}^{10^6} \varphi(d) = 303{,}963{,}552{,}391.
\]
This is computed efficiently with a \textbf{totient sieve} (analogous to the Sieve of Eratosthenes): initialise an array $\varphi[1..N]$ with $\varphi[d]=d$, then for each prime $p$ found, multiply $\varphi[kp]$ by $(1 - 1/p)$ for all multiples $kp$.
\end{example}

\subsubsection*{Counting Reduced Fractions in an Interval}

Sometimes we want reduced fractions $n/d$ in a strict interval $(a, b)$ rather than $(0,1)$.

\begin{example}[Fractions strictly between $1/3$ and $1/2$]
For each denominator $d$, the condition $1/3 < n/d < 1/2$ gives
\[
\frac{d}{3} < n < \frac{d}{2}
\implies n \in \left\{\left\lfloor\tfrac{d}{3}\right\rfloor + 1,\; \ldots,\; \left\lfloor\tfrac{d-1}{2}\right\rfloor\right\}.
\]
Among those numerators, count only those with $\gcd(n,d)=1$. For $d \le 12{,}000$ the total is $7{,}295{,}372$.

A direct double loop suffices:
\begin{verbatim}
long long ans = 0;
for (int d = 2; d <= 12000; d++) {
    int lo = d / 3 + 1, hi = (d - 1) / 2;
    for (int n = lo; n <= hi; n++)
        if (__gcd(n, d) == 1) ans++;
}
\end{verbatim}
\end{example}

\begin{remark}[Connection to Farey sequences]
The set of all reduced fractions in $[0,1]$ with denominator $\le N$ is the Farey sequence $F_N$. Consecutive fractions $a/b$ and $c/d$ in $F_N$ satisfy the \emph{mediant property}: $|bc - ad| = 1$.
\end{remark}

\subsection{2-Adic Valuation and the Lowbit Function}

\begin{definition}[2-adic valuation]
For a positive integer $n$, the \textbf{2-adic valuation} $v_2(n)$ is the exponent of the highest power of 2 dividing $n$:
\[
v_2(n) = \max\{k \in \mathbb{Z}_{\ge 0} : 2^k \mid n\}.
\]
\end{definition}

\begin{definition}[Lowbit]
The \textbf{lowbit} of $n$ is the value of the lowest set bit in the binary representation of $n$:
\[
\operatorname{lowbit}(n) = 2^{v_2(n)} = n \mathbin{\&} (-n).
\]
\end{definition}

\begin{example}
\begin{center}
\begin{tabular}{cccc}
$n$ & Binary & $v_2(n)$ & $\operatorname{lowbit}(n)$ \\
\hline
6 & $110_2$ & 1 & 2 \\
12 & $1100_2$ & 2 & 4 \\
8 & $1000_2$ & 3 & 8 \\
7 & $0111_2$ & 0 & 1 \\
\end{tabular}
\end{center}
When $n = 2^k$ is a power of 2, $\operatorname{lowbit}(n) = n$ itself and $v_2(n) = k$.
\end{example}

\begin{remark}[Bit-trick derivation]
In two's complement, $-n$ flips all bits of $n$ and adds 1. The lowest set bit of $n$ is the first 1 from the right; all bits below it are 0 in $n$ and 1 in $\sim n$. Adding 1 to $\sim n$ carries through those low-order 1s until it reaches the position of the original lowest set bit, which becomes 1 again. Hence $n \mathbin{\&} (-n)$ isolates exactly that lowest set bit.
\end{remark}

\subsubsection*{Sum of Lowbit Values}

\begin{theorem}[Recurrence for $T(N) = \sum_{m=1}^{N} \operatorname{lowbit}(m)$]
\[
T(2k) = k + 2\,T(k), \qquad T(2k+1) = T(2k) + 1.
\]
\end{theorem}
\begin{proof}
Among $\{1, 2, \ldots, 2k\}$: the $k$ odd numbers each have $\operatorname{lowbit} = 1$, contributing $k$. The $k$ even numbers satisfy $\operatorname{lowbit}(2j) = 2\,\operatorname{lowbit}(j)$, contributing $2T(k)$. So $T(2k) = k + 2T(k)$. Adding the odd number $2k+1$ contributes 1, giving $T(2k+1) = T(2k)+1$.
\end{proof}

\begin{remark}[Closed form at powers of 2]
By induction, $T(2^m) = (m+2)\cdot 2^{m-1}$.
\end{remark}

\subsection{Application: Hexagonal Grid Visibility}

A striking application of Euler's totient is counting hidden lattice points in a hexagonal grid.

\textbf{Setup.} Consider a hexagonal arrangement of points centred at the origin, where the $n$th order has points at distance up to $n$ in a regular hexagonal pattern. A point $P$ is \textbf{hidden} from the centre if another lattice point lies strictly between the origin and $P$ on the same ray.

\textbf{Reduction to one slice.} By 6-fold symmetry, count one $60^\circ$ slice and multiply by 6. In the slice, points form rows: row $k$ contains $k$ points.

\subsubsection*{Primitive Directions and Visible Points}

Label the $k$ points in row $k$ by the fractions
\[
\frac{0}{k},\; \frac{1}{k},\; \frac{2}{k},\; \ldots,\; \frac{k-1}{k}.
\]
This fraction encodes the direction (slope) of the ray from the origin to that point.

\begin{theorem}[Visible $\Leftrightarrow$ primitive direction]
The point labelled $a/k$ is \textbf{visible} from the origin if and only if $\gcd(a,k) = 1$.
\end{theorem}
\begin{proof}
If $\gcd(a,k) = d > 1$, then $a/k = (a/d)/(k/d)$, so the closer point in row $k/d$ lies on the same ray and blocks the view. Conversely, if $\gcd(a,k) = 1$, no closer point shares the same direction.
\end{proof}

\begin{corollary}
In row $k$: visible points $= \varphi(k)$; hidden points $= k - \varphi(k)$.
\end{corollary}

\subsubsection*{Formula for the Number of Hidden Points}

\begin{theorem}
The number of hidden points $H(n)$ in a hexagonal grid of order $n$ is
\[
H(n) = 6\sum_{k=1}^{n}\bigl(k - \varphi(k)\bigr) = 3n(n+1) - 6\,\Phi(n),
\]
where $\Phi(n) = \sum_{k=1}^{n}\varphi(k)$ is the summatory totient.
\end{theorem}

\begin{example}[$n = 5$]
$\varphi(1)=1,\;\varphi(2)=1,\;\varphi(3)=2,\;\varphi(4)=2,\;\varphi(5)=4$.
Hidden per row in one slice: $0, 1, 1, 2, 1$. Sum $= 5$. Multiply by 6: $H(5) = 30$.
\end{example}

\begin{remark}[Computational note for large $n$]
For $n = 10^8$, computing $H(n)$ requires $\Phi(n)$ efficiently. The summatory totient recursion (see above) with floor-grouping gives $\Phi(n)$ in $O(n^{2/3})$ time, making $H(10^8)$ tractable. The answer is $H(10^8) = 11762187201804552$.
\end{remark}
